\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-24 latex/article_vol2.tex
Chunks

\begin{Schunk}
\begin{Sinput}
R> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"    "integer"   "labelled"  "logical"  
[7] "numeric"  
\end{Soutput}
\end{Schunk}

@@

This function can now be called directly or used in \code{clean()}, as will
presented in an example below. Depending on the \code{doEval} argument, either a text string
with code or a plot is produced. The plot resulting from the following
call is found in Figure~\ref{fig:mosaicPlot}:
\begin{Schunk}
\begin{Sinput}
> mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)
\end{Sinput}
\end{Schunk}
% $

@@

As with \code{mosaicVisual()}, we change the class of this function in
order to make it appear in \code{allSummaryFunctions()} calls. But now we
wish to emphasize that the function is not intended to be called on
all variable types, as zeros have different roles in \code{Date}s and in
\code{logical} variables:
\begin{Schunk}
\begin{Sinput}
> countZeros <- summaryFunction(countZeros,
	description = "Count number of zeros",
	classes = c("character", "factor", "integer",
                    "labelled", "numeric"))
\end{Sinput}
\end{Schunk}
% \hl{more? don't end on code.}

\subsubsection{identifyColons - a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly:
\begin{Schunk}
\begin{Sinput}
identifyColons <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

  # Use regular expressions to identify colons between two words
  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

  if (length(problemValues) > 0) {
    problem <- TRUE
  }

  problemStatus <- list(problem = problem,
                        problemValues = problemValues)
  # Use the messagegenerator function to produce the output
  # message from the problemStatus and problemMessage
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)

identifyColons <- checkFunction(identifyColons,
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}
As with the previous two functions, we also change its class. Note,
however, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColon}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: %\hl{Jeg synes ikke testData egner sig her. Der er ingen koloner og CPRvar er et mærkeligt, internt navn, hvis ikke man kender danske CPR-numre. Langt de fleste variable har desuden ingen nuller. Måske skal der laves et nyt datasæt?}
\begin{Schunk}
\begin{Sinput}
> data("exampleData")
> clean(exampleData,
      # Add the new pre-check function
      preChecks = c("isKey", "isEmpty", "isID"),
      # Change the visual
      allVisuals = "mosaicVisual",
      # Add the new summary for specified data types
      characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
      factorSummaries = c(defaultFactorSummaries(), "countZeros"),
      labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
      numericSummaries = c(defaultNumericSummaries(), "countZeros"),
      integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
      # Add the new check for specific data types
      characterChecks = c(defaultCharacterChecks(), "identifyColons"),
      factorChecks = c(defaultFactorChecks(), "identifyColons"),
      labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}
The outputted document is found in Appendix \ref{sec:appendix2}.

@@

\subsubsection{Handling the figures}
Though figures give a nice overview of each variable, they are also
quite heavy objects in terms of memory allocation. Therefore, it might
be beneficial to not include figures in the \pkg{dataMaid} outputs for
very large datasets. This is controlled via the \code{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}
If figures are indeed needed, a different approach is to choose the
less memory heavy standard \proglang{R} figure style instead of the
\pkg{ggplot2} figures that are the default option in \code{clean()}. This
can be done by setting the \code{allVisuals = "basicVisual"} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}
Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot}.

@@

\begin{Schunk}
\begin{Sinput}
> clean(toyData, render = FALSE, openResult = FALSE)
> render("dataMaid_toyData.Rmd", quiet = FALSE)
\end{Sinput}
\end{Schunk}
This also deals with the fact that \pkg{dataMaid} can produce
\pkg{rmarkdown} files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the human time it takes to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. Or even for more reasonably sized
datasets, sometimes a brief overview of the most pressing issues can
be useful. This can be achieved by using the \code{onlyProblematic}
argument in \code{clean()}. By specifying \code{onlyProblematic = TRUE},
only variables that raise a flag in the checking steps will be
summarized and visualized. But perhaps we are not even interested in
obtaining general information about these variables, but only in
getting a quick overview of the problems they might have. This can be
done by also controlling the \code{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic = TRUE, mode = c("check"))
\end{Sinput}
\end{Schunk}
Now only the checking results are printed, and only for variables where problems were identified. An even more minimal output can be generated by also leaving out the checking results --- then \code{clean()} essentially just produces a list of the variable names that should be investigated further:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic = TRUE, mode = NULL)
\end{Sinput}
\end{Schunk}
Of course, this can also be done without generating an overview
document, by direct, interactive use of the \code{check()} function. When
called on a \code{data.frame}, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unravelled using the list manipulation function
\code{sapply()}:
\begin{Schunk}
\begin{Sinput}
> toyChecks <- check(toyData)
> foo <- function(x) {
>   any(sapply(x, function(y) y[["problem"]]))
> }
> sapply(toyChecks, foo)
\end{Sinput}
\begin{Soutput}
 var1  var2  var3  var4  var5  var6
 TRUE  TRUE  TRUE  TRUE  TRUE FALSE
\end{Soutput}
\end{Schunk}
and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free.

@@

In the this example, the \code{dataMaid\_toyData.Rmd} file could have been created as follows:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, standAlone = FALSE)
\end{Sinput}
\end{Schunk}
and the more minimal, html-style \pkg{rmarkdown} file described above can be produced using
\begin{Schunk}
\begin{Sinput}
> clean(toyData, standAlone = FALSE, output = "html", twoCols = FALSE)
\end{Sinput}
\end{Schunk}
Note that with the latter option, no special YAML preamble is needed in the \pkg{rmarkdown} document.
% \hl{don't end on code.}

>>>>

\begin{Schunk}
\begin{Sinput}
R> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"    "integer"   "labelled" 
[6] "logical"   "numeric"  
\end{Soutput}
\end{Schunk}

@@

This function can now be called directly or used in \code{clean()}, as will
presented in an example below. Depending on the \code{doEval} argument, either a text string
with code or a plot is produced. The plot resulting from the following
call is found in Figure~\ref{fig:mosaicPlot}:

\begin{Schunk}
\begin{Sinput}
R> mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)
\end{Sinput}
\end{Schunk}

@@

As with \code{mosaicVisual()}, we change the class of this function in
order to make it appear in \code{allSummaryFunctions()} calls. But now we
wish to emphasize that the function is not intended to be called on
all variable types, as zeros have different roles in \code{Date}s and in
\code{logical} variables:

XXX indentation
\begin{Schunk}
\begin{Sinput}
R> countZeros <- summaryFunction(countZeros,
+    description = "Count number of zeros",
+    classes = c("character", "factor", "integer",
+    "labelled", "numeric"))
\end{Sinput}
\end{Schunk}

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly

XXX comments

\begin{Schunk}
\begin{Sinput}
R> identifyColons <- function(v, nMax = Inf, ... ) {
+    v <- unique(na.omit(v))
+    problemMessage <- "Note: The following values include colons:"
+    problem <- FALSE
+    problemValues <- NULL
+  
+    # Use regular expressions to identify colons between two words
+    problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
+                              function(x) all(x != -1))]
+  
+    if (length(problemValues) > 0) {
+      problem <- TRUE
+    }
+  
+    problemStatus <- list(problem = problem,
+                          problemValues = problemValues)
+    # Use the messagegenerator function to produce the output
+    # message from the problemStatus and problemMessage
+    outMessage <- messageGenerator(problemStatus, problemMessage, nMax)
+  
+    checkResult(list(problem = problem,
+                     message = outMessage,
+                     problemValues = problemValues))
+  }
R> identifyColons <- checkFunction(identifyColons,
+      description = "Identify colons surrounded by alphanumeric characters",
+      classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}

As with the previous two functions, we also change its class. Note,
however, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColon}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: %\hl{Jeg synes ikke testData
                                %egner sig her. Der er ingen koloner
                                %og CPRvar er et mærkeligt, internt
                                %navn, hvis ikke man kender danske
                                %CPR-numre. Langt de fleste variable
                                %har desuden ingen nuller. Måske skal
                                %der laves et nyt datasæt?}
\begin{Schunk}
\begin{Sinput}
R> data("exampleData")
R> clean(exampleData,
+    preChecks = c("isKey", "isEmpty", "isID"),
+    allVisuals = "mosaicVisual",
+    characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
+    factorSummaries = c(defaultFactorSummaries(), "countZeros"),
+    labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
+    numericSummaries = c(defaultNumericSummaries(), "countZeros"),
+    integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
+    characterChecks = c(defaultCharacterChecks(), "identifyColons"),
+    factorChecks = c(defaultFactorChecks(), "identifyColons"),
+    labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}

The outputted document is found in Appendix \ref{sec:appendix2}.

@@

\subsubsection{Handling the figures}
Though figures give a nice overview of each variable, they are also
quite heavy objects in terms of memory allocation. Therefore, it might
be beneficial to not include figures in the \pkg{dataMaid} outputs for
very large datasets. This is controlled via the \code{mode} argument:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}

If figures are indeed needed, a different approach is to choose the
less memory heavy standard \proglang{R} figure style instead of the
\pkg{ggplot2} figures that are the default option in \code{clean()}. This
can be done by setting the \code{allVisuals = "basicVisual"} argument:

Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot} \citep{txtplot}.

@@

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, render = FALSE, openResult = FALSE)
R> render("dataMaid_toyData.Rmd", quiet = FALSE)
\end{Sinput}
\end{Schunk}

This also deals with the fact that \pkg{dataMaid} can produce
\pkg{rmarkdown} files that supersedes the upper size limit of code editors, e.g.,
RStudio, which currently has an editor file size limit of 5MB. %\hl{dette giver
%ingen mening uden en præcision? "upper size limit" af hvad? Se udkommenteret
%kommentar.}
% of RStudio,
%which is currently \hl{find number} GBs (using RStudio version
%1.0.44).
%\hl{Is this maybe too editor specific? On the other hand, a
%  lot of people do use RStudio...}.

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the data is large, but memory issues and computation time are less
of an issue than the human time it takes to look through the data
cleaning document, a viable solution might be not to include all
information about all variables. Or even for more reasonably sized
datasets, sometimes a brief overview of the most pressing issues can
be useful. This can be achieved by using the \code{onlyProblematic}
argument in \code{clean()}. By specifying \code{onlyProblematic = TRUE},
only variables that raise a flag in the checking steps will be
summarized and visualized. But perhaps we are not even interested in
obtaining general information about these variables, but only in
getting a quick overview of the problems they might have. This can be
done by also controlling the \code{mode} argument:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, onlyProblematic = TRUE, mode = c("check"))
\end{Sinput}
\end{Schunk}

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output can be
generated by also leaving out the checking results --- then
\code{clean()} essentially just produces a list of the variable names
that should be investigated further:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, onlyProblematic = TRUE, mode = NULL)
\end{Sinput}
\end{Schunk}

Of course, this can also be done without generating an overview
document, by direct, interactive use of the \code{check()} function. When
called on a \code{data.frame}, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unravelled using the list manipulation function
\code{sapply()}:

\begin{Schunk}
\begin{Sinput}
R> toyChecks <- check(toyData)
R> foo <- function(x) {
+    any(sapply(x, function(y) y[["problem"]]))
+  }
R> sapply(toyChecks, foo)
\end{Sinput}
\begin{Soutput}
 var1  var2  var3  var4  var5  var6 
 TRUE  TRUE  TRUE  TRUE  TRUE FALSE 
\end{Soutput}
\end{Schunk}

and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free.

@@

In the this example, the \code{dataMaid\_toyData.Rmd} file could have
been created as follows:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, standAlone = FALSE)
\end{Sinput}
\end{Schunk}

and the more minimal, html-style \pkg{rmarkdown} file described above can be produced using

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, standAlone = FALSE, output = "html", twoCols = FALSE)
\end{Sinput}
\end{Schunk}

Note that with the latter option, no special YAML preamble is needed in the \pkg{rmarkdown} document.
% \hl{don't end on code.}
